---
title: "STAT 184 Final Project"
author: "Gavin Crowley and Alex Bechakas"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi = 300)

# Load all packages

library(readxl)
library(tidyr)
library(dplyr)
library(ggplot2)
library(kableExtra)

```

(All code should be written according to a Style Guide of your choice. List this Style Guide as a code comment in your first code chunk)

# Topic

For our project, we are using two NFL Data sets. We chose to have our project be about NFL data because we are both sports fans and specifically big NFL fans and we are both interested in the overall breakdown of statistics in football.

# Investigation

## For the defense data, we are investigating:
How each statistic varies by position?
What positions lead in what stats?

## For the Quarterback data, we are investigating: 
How do Quarterbacks stats compare between eras?
What Quarterback had the best season?

# Defense Data Set 

This data set is from Pro Football Reference. The purpose of this data set is that it contains the defensive statistics for all players who recorded at least 1 defensive stat in the 2023/24 NFL Season. A case in this data set is a individual player. Our analysis of the defense data set will focus on most of the attributes. Those attributes include: Position, G, GS, Int, TD, PD, FF, FR, Sk, Comb, Solo, Ast, TFL, QBHits, and Sfty.

```{r}
DefenseStats <- read_excel("Penn State/STAT 184/Final Project/FP_GavinCrowley_AlexBechakas/2023 NFL Defense Stats.xlsx")
View(DefenseStats)
```


# Defense Data Set Wrangling

Selected the columns from the data set we were interested in using excluding columns like player name, team, and return yards as those were not needed.
Filtered the position column to only include defensive positions and not include offensive positions.
Grouped all the data by position so the table would have one row for each position and the entries in each column would be the total stats for all players in that position.

```{r}
DefS <- DefenseStats %>%
  select(Position,G,GS,Int,TD,PD,FF,FR,Sk,Comb,Solo,Ast,TFL,QBHits, Sfty)
DefS <- DefS %>% filter (Position %in% c("DT", "DE", "LB", "CB", "S"))
DefS <- DefS %>% group_by(Position) %>%
  summarize(
    Games = sum(G),
    Games_Started = sum(GS),
    Interceptions = sum(Int),
    Touchdowns = sum(TD),
    Pass_Deflections = sum(PD),
    Forced_Fumbles = sum(FF),
    Fumble_Recoveries = sum(FR),
    Sacks = sum(Sk),
    Combined_Tackles = sum(Comb),
    Solo_Tackles = sum(Solo),
    Assisted_Tackles = sum(Ast),
    Tackles_For_Loss = sum(TFL),
    QB_Hits = sum(QBHits),
    Safeties = sum(Sfty), .groups = "drop"

  )
```


# Defense Data Table

```{r}
DefS %>%
  kable(
    caption = "Statistic Totals for Each NFL Position",  
    booktabs = TRUE,                        
    align = rep("c", 11)           
  ) %>%
  kable_styling(      
    bootstrap_options = c("striped"),     
    font_size = 16
  )
```
For this data table, each row is a different position. Each column is a different stat and the entries are the total of that recorded stat for all players of that position.

# Defense Data Visualizations

For the Defense Data, we created 14 data visualizations. Each data visualization has position as the x-axis and the stat as the y-axis. Each data visualization shows which position leads in that statistic.

## Games Graph

```{r}
ggplot(DefS) +
  aes(x = Position, y = Games) +
  geom_col(fill = "#212221") +
  labs(
    x = "Position",
    y = "Games",
    title = "Total Games by Position for the 2023/2024 NFL Season"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```

This graph shows that linebackers lead in games played and defensive ends have the least games played.

## Games Started Graph

```{r}
ggplot(DefS) +
  aes(x = Position, y = Games_Started) +
  geom_col(fill = "#212221") +
  labs(
    x = "Position",
    y = "Games Started",
    title = "Total Games Started by Position for the 2023/2024 NFL Season"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )

```

This graph shows that linebackers lead in games started and defensive ends have the least games started.

## Interceptions Graph

```{r}
ggplot(DefS) +
  aes(x = Position, y = Interceptions) +
  geom_col(fill = "#212221") +
  labs(
    x = "Position",
    y = "Interceptions",
    title = "Total Interceptions by Position for the 2023/2024 NFL Season"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```

This graph shows that corner backs lead in interceptions and defensive ends have the least interceptions.

## Touchdowns Graph

```{r}
ggplot(DefS) +
  aes(x = Position, y = Touchdowns) +
  geom_col(fill = "#212221") +
  labs(
    x = "Position",
    y = "Touchdowns",
    title = "Total Touchdowns by Position for the 2023/2024 NFL Season"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```

This graph shows that corner backs lead in touchdowns and defensive ends have the least touchdowns.

## Pass Deflections Graph

```{r}
ggplot(DefS) +
  aes(x = Position, y = Pass_Deflections) +
  geom_col(fill = "#212221") +
  labs(
    x = "Position",
    y = "Pass Deflections",
    title = "Total Pass Deflections by Position for the 2023/2024 NFL Season"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```

This graph shows that corner backs lead in pass deflections and defensive ends have the least pass deflections.

## Forced Fumbles Graph

```{r}
ggplot(DefS) +
  aes(x = Position, y = Forced_Fumbles) +
  geom_col(fill = "#212221") +
  labs(
    x = "Position",
    y = "Forced Fumbles",
    title = "Total Forced Fumbles by Position for the 2023/2024 NFL Season"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```

This graph shows that linebackers lead in forced fumbles and defensive ends have the least forced fumbles.

## Fumble Recoveries Graph

```{r}
ggplot(DefS) +
  aes(x = Position, y = Fumble_Recoveries) +
  geom_col(fill = "#212221") +
  labs(
    x = "Position",
    y = "Fumble Recoveries",
    title = "Total Fumble Recoveries by Position for the 2023/2024 NFL Season"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```

This graph shows that linebackers lead in fumble recoveries and defensive ends have the least fumble recoveries.

## Sacks Graph

```{r}
ggplot(DefS) +
  aes(x = Position, y = Sacks) +
  geom_col(fill = "#212221") +
  labs(
    x = "Position",
    y = "Sacks",
    title = "Total Sacks by Position for the 2023/2024 NFL Season"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```

This graph shows that linebackers lead in sacks and corner backs have the least sacks.

## Combined Tackles Graph

```{r}
ggplot(DefS) +
  aes(x = Position, y = Combined_Tackles) +
  geom_col(fill = "#212221") +
  labs(
    x = "Position",
    y = "Combined Tackles",
    title = "Total Combined Tackles by Position for the 2023/2024 NFL Season"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```

This graph shows that linebackers lead in combined tackles and defensive ends have the least combined tackles.

## Solo Tackles Graph

```{r}
ggplot(DefS) +
  aes(x = Position, y = Solo_Tackles) +
  geom_col(fill = "#212221") +
  labs(
    x = "Position",
    y = "Solo Tackles",
    title = "Total Solo Tackles by Position for the 2023/2024 NFL Season"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```

This graph shows that linebackers lead in solo tackles and defensive ends have the least solo tackles.

## Assisted Tackles Graph

```{r}
ggplot(DefS) +
  aes(x = Position, y = Assisted_Tackles) +
  geom_col(fill = "#212221") +
  labs(
    x = "Position",
    y = "Assisted Tackles",
    title = "Total Assisted Tackles by Position for the 2023/2024 NFL Season"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```

This graph shows that linebackers lead in assisted tackles and defensive ends have the least assisted tackles.

## Tackles For Loss Graph

```{r}
ggplot(DefS) +
  aes(x = Position, y = Tackles_For_Loss) +
  geom_col(fill = "#212221") +
  labs(
    x = "Position",
    y = "Tackles For Loss",
    title = "Total Tackles For Loss by Position for the 2023/2024 NFL Season"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```

This graph shows that blank lead in blank and blank have the least blank

## QB Hits Graph

```{r}
ggplot(DefS) +
  aes(x = Position, y = QB_Hits) +
  geom_col(fill = "#212221") +
  labs(
    x = "Position",
    y = "QB Hits",
    title = "Total QB Hits by Position for the 2023/2024 NFL Season"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```

This graph shows that linebackers lead in quarter back hits and corner backs have the least quarter back hits.

## Safeties Graph

```{r}
ggplot(DefS) +
  aes(x = Position, y = Safeties) +
  geom_col(fill = "#212221") +
  labs(
    x = "Position",
    y = "Safeties",
    title = "Total Safeties by Position for the 2023/2024 NFL Season"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )
```

This graph shows that defensive tackles lead in safeties and safeties have the least safeties.

Overall, linebackers lead in most of the statistics and defensive ends have the least of most statistics. This makes sense as linebackers have the most games played and started and defensive ends have the least.

# Quarterback Data Set

This data set is from kaggle.


# Quarterback Data Set Wrangling



# Quarterback Data Table



# Quarterback Data Visualizations


